---
title: 'Visualization in Base R'
author: "Jamison R. Crawford, MPA"
date: "March 13, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}

options(scipen = 999)

if(!require(zoo)){install.packages("zoo")}
if(!require(tidyr)){install.packages("tidyr")}
if(!require(readr)){install.packages("readr")}
if(!require(dplyr)){install.packages("dplyr")}
if(!require(GGally)){install.packages("GGally")}
if(!require(scales)){install.packages("scales")}
if(!require(stringr)){install.packages("stringr")}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(kableExtra)){install.packages("kableExtra")}

library(zoo)
library(tidyr)
library(readr)
library(dplyr)
library(GGally)
library(scales)
library(stringr)
library(ggplot2)
library(kableExtra)

rm(list = ls())

```

<br>

# What is Data Visualization?

<br>

**Data Visualization** is:

* "The representation of an object, situation, or set of information as a chart or other image" [- Oxford Dictionary](https://en.oxforddictionaries.com/definition/us/data_visualization)
* "...the creation and study of the visual representation of data" [- Wikipedia](https://en.wikipedia.org/wiki/Data_visualization)

<br>

**Abbreviations****:** Both **"data vis"** and **"data viz"** are used as abbreviations for data visualization.

<br>

Observe this example, which visualizes public construction records for Lakeview Amphitheater in Syracuse, NY:

<br>

<center>

```{r, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
url <- "https://raw.githubusercontent.com/jamisoncrawford/reis/master/Datasets/tblr_master.csv"
lvhc <- read_csv(url) %>%
  filter(!is.na(race))

index <- which(lvhc$race == "Native")
lvhc[index, "race"] <- "Indigenous"

ggplot(lvhc, aes(x = factor(race, levels = rev(c("White",
                                             "Indigenous",
                                             "Hispanic",
                                             "Black",
                                             "Multiracial",
                                             "Asian"))), y = gross)) +
  geom_jitter(stat = "identity", width = 0.3, alpha = 0.4, color = "tomato") +
  coord_flip() +
  theme_minimal() +
  labs(x = NULL,
       y = "Gross (K)",
       title = "An example of data visualization",
       subtitle = "Lakeview Amphitheater: Gross construction worker payments by race",
       caption = "Source: Onondaga County\nRecords disclosing race") +
  scale_y_continuous(labels = c("$ 0", "1", "2", "3", "4 K")) +
  theme(axis.title.x = element_text(vjust = -0.35,
                                    color = "grey40"),
        axis.text.y = element_text(color = "black"),
        axis.text.x = element_text(color = "black"),
        plot.caption = element_text(vjust = -0.25,
                                    color = "grey60",
                                    lineheight = 1.2),
        plot.subtitle = element_text(color = "grey 40"))

```

</center>

<br>

**Questions****:** In observing the above scatter plot:

* What do you like about it?
* What do you dislike about it?
* What idea does this image attempt to convey? 

<br>

# Visualization in R

<br>

**R** uses, or may use, both built-in and external **packages** (libraries) to produce graphics.

The three most popular packages, from oldest to newest, are:

* [graphics](https://www.rdocumentation.org/packages/graphics/versions/3.5.2), built into R for "base graphics"
* [lattice](https://cran.r-project.org/web/packages/lattice/index.html), an external package for [small multiples](https://en.wikipedia.org/wiki/Small_multiple), which group and depict data in small charts on common scales
* [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html), a core [Tidyverse](https://www.tidyverse.org/) package based on Leland Wilkinson's "Grammar of Graphics" (1999) 

<br>

Honorable mention goes to package [grid](https://www.rdocumentation.org/packages/grid/versions/3.5.2), which powers a lot of base R `graphics` output behind the scenes.

<br>

# Base R "graphics"

<br>

Package `graphics` comes preinstalled and preloaded in your *R* or *RStudio* environment.

* Base R visualizations are layered, i.e. an "Artist's Palette" approach
* The first visualization activates R's graphics device
* Subsequent commands layer over the initial viz
* Errors require rerunning these commands

<br>

## Practice Data

<br>

**Let's import** the Lakeview Amphitheater data again and save them in object `lv`:

```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
library(readr)

url <- "https://raw.githubusercontent.com/jamisoncrawford/reis/master/Datasets/tblr_master.csv"
lv <- read_csv(url)

index <- which(lv$project == "Lakeview")
lv <- lv[index, 1:12]
```

You can find the documentation for these data in the [Lakeview](https://github.com/jamisoncrawford/lakeview) and [Hancock](https://github.com/jamisoncrawford/hancock) repositories.

<br>

**Explore the data** with `str()` and other exploratory functions:

```{r, eval=FALSE}
names(lv)
dims(lv)
summary(lv)
```

```{r}
str(lv)
```

<br>

## Functions

<br>

**Plotting Functions****:** Base R graphics have many functions. The following will initialize a plot:

* `plot()` automatically detects/creates scatter plots, bar charts, etc.
* `hist()` creates a histogram for a single variable
* `boxplot()` creates a box and whisker plot
* `barplot()` creates bar plots

There are several other functions for intializing plots, but these are common.

<br>

**Annotating Functions****:** The functions and others will annotate an existing plot.

* `loess()` adds a smoother to an existing scatter plot
* `text()` adds text to an existing plot
* `jitter()` adds random noise to points to break up overlap
* `legend()` adds a legend to an existing plot

<br>

**Modifiable Parameters****:** Parameters go inside the `()` of a function, and there are many.

* `main =` modifies title
* `xlab =` modifies x-axis label
* `ylab =` modifies y-axis label
* `lwd =` modifies line width
* `lty =` modifies line type
* `col =` modifies color
* `cex =` modifies size

There are a ridiculous number of modifiable parameters. You can learn more by running `?par`.

The curious learner should check out this [Base R Cheat Sheet](https://dcgerard.github.io/stat234/base_r_cheatsheet.html) by David Gerard (2017).

<br>

**Turn It Off****:** Since each annotating function adds a layer to an existing plot:

* You can begin a new plot with `plot.new()`
* You can turn off (and restart) your graphical device with `dev.off()`

These are great go-to procedures for troubleshooting when visualizations subvert from expectations.

<br>

**Let's Plot****:** The following is a scatterplot of `hours` worked as a function of `gross` pay.

* Here, we're setting `x =` and `y=` to dataset `lv` and variables `hours` and `gross` using `$`

```{r}
plot.new()
plot(x = lv$hours, y = lv$gross)
```

<br>

**Modifying an Initial Plot****:** You can adjust the parameters in the initial call to `plot()`.

```{r}
plot.new()
plot(x = lv$hours, 
     y = lv$gross,
     main = "Lakeview: Weekly hours v. gross pay",
     xlab = "Hours",
     ylab = "Gross")
```

<br>

**Experiment** with the above plot with new arguments. Try:

* `pch =` adjusts point shape (e.g. `pch = 19`)
* `col =` adjust color (e.g. `col = "tomato", try it)

<br>

## Mid-Introduction Practice

<br>

**Instructions****:** Use these data (`lv`) to create a histogram, function `hist()`, of the distribution of hours worked each pay period, variable `hours`. Refer to these data as `lv$hours` in arguments.

The following arguments are separated by commas and must use `""`, e.g. `col = "skyblue"`

Check out [Colors in R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) for a sizeable selection of colors.

* Set a title with argument `main =`
* Set a color argument with `col =`
* Set an x-axis label with `xlab =`
* Set a y-axis label with `ylab =`
* Set the size of the binwidths with `breaks =`
* Store this in object: `my_hist` - the code is all set

```{r eval=FALSE}
my_hist <- hist()
```

<br>

## Global Parameters

<br>

**Setting Global Parameters****:** Recall the earlier call to `?par` to view all base graphics arguments.

* Function `par()` allows setting global parameters
* In other words, they apply to plots until reset
* Allows for unified and consistent visualizations
* Reset by calling `par()` or `dev.off()`

<br>

For example, we can change the number of visualizations in one graphic with argument `mfrow =`:

```{r eval=FALSE}
par(mfrow = c(1, 2))
```

Here, `mfrow =` takes a pair of numbers in function `c()`, the first is for rows, the second, columns.

<br>

**Grid It****:** After calling `par()` with `mfrow = c(1, 2)`, you can call the following plot functions:

```{r}
par(mfrow = c(1, 2))

plot(x = lv$hours, 
     y = lv$gross,
     main = "Weekly hours v. gross",
     xlab = "Hours",
     ylab = "Gross (USD)",
     pch = 21,
     col = "tomato")

hist(x = lv$hours,
     main = "Distribution of worker hours",
     xlab = "Hours",
     col = "skyblue")
```

<br>

## Exporting Graphics

<br>

**Base exporting** functions include `png()`, `pdf()`, `jpeg()`, `tiff()`, `bmp()`, and others.

* `filename =` takes a quoted value, including an extension, e.g. `"hours_distro.png"`
* `height =` and `width = ` modify dimensions - important for tailoring, not stretching, graphics
* `units =`, relating to height and width, is the units for size, e.g. `"in"` (inches) or `"px"` (pixels)
* `family =` selects font family, affecting all text
* `res =` adjusts resolution

<br>

## Verdict

<br>

**Weaknessess**:** Base R package `graphics` has shortcomings:

* "Artist's Palette" is cool, and good script writing helps, but it may be inconvenient
* Limited use of [small multiples](https://en.wikipedia.org/wiki/Small_multiple) with function `pairs()`
* Inconsistent in syntax, with unintuituve arguments (e.g. `cex =`)
* Limited in displaying multivariate data
* Not terribly elegant

<br>

**Strengths****:** Base R package `graphics` has a few advantages:

* "Quick and Dirty" exploratory visualizations to understand new data
* Uses many arguments which help in understanding more advanced packages
* Pretty good for statistics-related visualization

<br>

**In summary,** Base R `graphics` is rarely used when alternatives like package `ggplot2` are easier to learn and often more practical to implement. However, it has some pretty nifty statistical uses. Here's a few:

**Heatmaps**

```{r}
data(volcano)
heatmap(volcano)
```

**Model Summaries**

```{r}
my_lm <- lm(gross ~ hours, data = lv)
par(mfrow = c(2, 2))
plot(my_lm)
```

</center>

<br>

**Dendrograms**

```{r, cache=TRUE}
my_cluster <- hclust(dist(USArrests), "ave")
plot(my_cluster)
```

<br>

**Mosaic Plots or Confusion Matrices**

```{r}
mosaicplot(gross ~ hours, 
           data = lv[1:20, ])
```

<br>

**Univariate Diagnostics**



<br>

# Applied Practice

<br>

**Importing Data****:** Run the following code to import the worker records for the I-690.

```{r warning=FALSE, message=FALSE}
library(readr)

url <- "https://raw.githubusercontent.com/jamisoncrawford/reis/master/Datasets/690_workforce_summary.csv"
hw <- read_csv(url)[, 1:12]
```

 <br>

**The Plots****:** Create four plots in a 2 by 2 grid. The calls to `par()` and `mfrow =` are in the code below. Create:

1. A histogram of `hw$hours` with function `hist()`
2. A histogram of `hw$gross` with function `hist()`
3. A scatterplot of `hw$gross` and `hw$hours` with function `plot()`
4. A boxplot (using function `plot()`) with `x = as.factor(hw$race)` and `y = hw$gross`

<br>

**The Annotations****:** Include the following arguments for each plot and supply thoughtful parameters:

* `main =`
* `xlab =` (optional)
* `ylab =` (optional)
* Set `frame.plot = FALSE`
* Use `col =`, `cex =`, `size =`, and `pch =` at your discretion

```{r evaluate=FALSE}
par(mfrow = c(2, 2))
```