---
title: 'Visualization with ggplot2: Part II'
author: "Jamison R. Crawford, MPA"
date: "April 26, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---

<style>
body {
text-align: justify}
</style>

<br>

# Review

**Base R Visualization.** For an introduction to data visualization in R, see [Visualization in Base R](http://rpubs.com/JamisonCrawford/graphics).

**Introduction to ggplot2.** To view the first part of this series on visualization in `ggplot2`, see [Visualization with ggplot2: Part I](http://rpubs.com/jamisoncrawford/ggplot2).

<br>

## Installing & Loading

**Installing Packages.** Packages, or libraries, extend R's functionality.

  * Install packages with function `install.packages()` and the package name in quotes `""`
  * If you already have `ggplot2` installed, you do not need to reinstall it, just load it
  * The following command will install package `ggplot2`
  
<br>
    
```{r message=FALSE, warning=FALSE, eval=FALSE}
install.packages("ggplot2")
```

<br>

**Additional Packages.** Because we'll focus on graphical themes in this session:

  * We'll also install package `ggthemes` and `GGally` to enhance the `ggplot2` package
  * Like `ggplot2`, skip this if `ggthemes` is already installed
  * Package `ggthemes` adds a number of new premade themes to `ggplot2`
  * Explore dozens of new themes at [All Your Figure Are Belong to Us](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)
  * Remember that R is *case-sensitive*
  
<br>
    
```{r message=FALSE, warning=FALSE, eval=FALSE}
install.packages("ggthemes")
install.packages("GGally")
```

<br>

**Loading Packages.** Once installed, packages must be loaded into your session:

  * You can load packages individually using function `library()`
  * This time, quotes (`""`) are optional for package names
  * Remember that R is *case-sensitive*
  
<br>

```{r message=FALSE, warning=FALSE, eval=TRUE, cache=TRUE}
library(ggplot2)
library(ggthemes)
library(GGally)
```

<br>

## The Grammar of Graphics

**Visualization Shorthand.** Like human language and parts of speech, e.g. verbs, nouns, adjectives, etc.: 

  * Data visualization is also comprised of "parts" called **layers**
  * See Leland Wilkinson's work [The Grammar of Graphics](https://books.google.com/books/about/The_Grammar_of_Graphics.html?id=NRyGnjeNKJIC&printsec=frontcover&source=kp_read_button#v=onepage&q&f=false) (1999)
  * "Grammar of Graphics" is the `gg` in `ggplot2`
  * `ggplot2` was designed by Hadley Wickham using this framework
  * See Wickham's [A Layered Grammar of Graphics](https://vita.had.co.nz/papers/layered-grammar.html) (2010)
  
<br>
  
**Layers.** "Layers" are the equivalent of parts of speech in human language; there are 7 in total:

  * **Data.** Your dataset and variables of interest (`ggplot()`)
  * **Aesthetics.** How you map variables, e.g. x- and y-axis, color, size, etc. (`aes()`)
  * **Geometries.** The shape of your visualization, e.g. scatter plots, bar plots, etc. (`geom_*()`)
  * **Facets.** Grouping variables and displaying each over many visualizations (`facet_*()`)
  * **Statistics.** Statistical elements, such as bins and linear regression (`stat_*()`)
  * **Coordinates.** Zooming in or changing height and width (`coord_*()`)
  * **Themes.** Stylizing and polishing plots (`theme_*()`)
  
<br>

**Essential Layers.** Three **layers** are essential to any `ggplot2` visualization:

  * **Data.** Because you need some data to visualize them
  * **Aesthetics.** Because you want to show specific variables from your data
  * **Geometries.** Because your variables must assume a form
  
<br>

**Observe.** Here we use the practice dataset `diamonds`, which we call with function `data()`:

  * Note that we connect `ggplot2` functions with the **addition operator**, or `+`
  * The `diamonds` dataset is built into `ggplot2`
  * Optional: Load `diamonds` with function `data()`
  
<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds) +                # Data
  aes(x = carat, y = price) +     # Aesthetics, here x- and y-axes
  geom_point()                    # Geometries, here a scatterplot
```
  
<br>

**"Proper" Syntax.** Though the above is easier, function `aes()` *should be* nested in `ggplot()`

```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds,             # Separate arguments with commas
       aes(x = carat, 
           y = price)) +     # Double end parantheses ends both ggplot() and aes()
  geom_point()
```
  
<br>

## Data vs. Non-Data Ink

**Data Ink.** The **aesthetics layer** is only used to show **data ink**, in function `aes()`

  * "Data Ink" refers to any element of a page that depicts your data
  * "Data Ink" is dynamic, so when your data changes, your ink changes
  * Arguments include `x =`, `y =`, `color =`, `fill =`, `size =`, etc.
  * Also called **aesthetic mappings**
  
<br>
  
```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds,
       aes(x = carat, 
           y = price,
           color = clarity)) +     # Argument "color =" shows variable "clarity"
  geom_point()
```
  
<br>

**Non-Data Ink.** The **geometries layer** also shows **non-data ink** in `geom_*()` functions

  * "Non-data Ink" does not depict data, but makes graphics easier to interpret
  * "Non-data Ink" is not dynamic, so it stays the same if your data change
  * Except for `x =` and `y =`, all **data ink** arguments are the same
  * Also called **attributes**
  
<br>
  
```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds,
       aes(x = carat, 
           y = price)) +
  geom_point(alpha = 0.075,        # Argument "alpha =" makes data transparent (7.5%)
             color = "tomato")     # Argument "color =" makes your data pop
```

<br>

**Conclusion.** Combine both **data** and **non-data ink** for optimal effect.

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds,
       aes(x = carat, 
           y = price,
           color = clarity)) +
  geom_point(alpha = 0.35)        # Argument "alpha =" makes data transparent (7.5%)
```

<br>

# Geometries

There are 60+ geometry functions in package `ggplot2`. They all begin with `geom_`.

  * To browse geometries, simply type `geom_` in the RStudio **console**
  * Alternatively, you can view the [official ggplot2 documentation](https://ggplot2.tidyverse.org/reference/)
  * Alternatively, you can view the [official ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) 

<br>

## Proper Form

**What's Good?** Over time, which **geometries** to use becomes intuitive.

  * Selecting the right **geometry** begins with one question: "What's the big idea?"
  * If visualizing for a colleague or collaborator, always *consult for context*
    - "I want to show revenue over time" indicates a **line chart**
    - "I want to show the difference in income by ethnicity" indicates a **bar chart**
    - "I want to that most of our clients are young folks" indicates a **histogram**
    - "I want to show a pie chart with all 23 flavors in Dr. Pepper" indicates **inexperience**
  * If you need help, consider visiting "Input" in [Data Viz Project](https://datavizproject.com/)

<br>

**Scatter Plots** measure two continuous (numeric) variables, like length and width:

  * Use function `geom_point()` for a typical scatter plot
  * If many overlapping points, add random noise with `geom_jitter()`

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
data(iris)
ggplot(iris, 
       aes(x = Petal.Length,       # Length in centimeters
           y = Petal.Width,        # Width in centimeters
           color = Species)) +
  geom_point()

ggplot(iris, 
       aes(x = Petal.Length,       # Length in centimeters
           y = Petal.Width,        # Width in centimeters
           color = Species)) +
  geom_jitter()
```

<br>

**Bar Plots** measure one continuous (numeric) and one categorical variable:

  * Note that `geom_bar()` needs argument `stat =` to tell R how to plot your geometry
  * Calling `options(scipen = 999)` disables scientific notation (orders of magnitude)

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
options(scipen = 999)             # Disable scientific notation
```

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
ggplot(diamonds, 
       aes(x = cut,               # Categorical (Quality of Cut)
           y = price)) +          # Continuous (US Dollars)
  geom_bar(stat = "identity",     # Argument "stat =" tells R how to handle stats
           color = "skyblue")     
```

<br>

**Histograms** visualize distributions of a single continuous variable, like `mpg`:

  * Function `geom_histogram()` automatically chooses **bin widths**
  * **Histograms** show total amount of observations in each range of values
  * **Bin widths** control how big or small each value range is
  * Note that `fill =` is needed for areas and `color =` is for lines and points 

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
data(diamonds)
ggplot(diamonds,
       aes(x = carat)) +
  geom_histogram(fill = "dodgerblue3")     # Argument "fill =" needed for areas
```

<br>

**Line Graphs** generally visualize changes in data over time.

  * There are several types of line graphs in `ggplot2`
    - `geom_line()` seems obvious, but it makes the line touch the x-axis between values
    - `geom_step()` is an improvement, making horizontal lines between values
    - `geom_path()` is often ideal, directly connecting each data point with a line
  * Practice dataset `economics` already has dates and number of unemployed, *in thousands*

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
data(economics)
ggplot(economics, 
       aes(x = date,
           y = unemploy)) +
  geom_path()
```

<br>

**Conclusion.** Choose the right **geometries** for the variables you want to show. 

*Then* make it pretty. But never *too pretty* that it obfuscates the clarity of your ideas.

<br>

> "His clothes were all new and had been cut by a good Moscow tailor. But there was something wrong even with his clothes. They were rather too fashionable, as clothes always are from conscientious but not very talented tailors." 

*Fyodor Dostoevsky, "The Idiot", Part II, Chapter 2*

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE, echo=FALSE}
library(scales)
ggplot(economics, 
       aes(x = date,
           y = unemploy)) +
  geom_path() +
  labs(title = "National unemployment",
       subtitle = "United States, 1967-2015",
       caption = "Source: Federal Reserve Bank of St. Louis",
       x = "Year",
       y = "Unemployed, in thousands") +
  scale_y_continuous(labels = comma, 
                     limits = c(0, 16000)) +
  theme_minimal()
```

<br>

## Factor This In

**Factors** in R are the same as **categorical variables**, a.k.a. "nominal", "discrete", etc.

  * This is a source of frustration, since you sometimes must tell R explicitly
  * Use function `as.factor()` on a variable to tell R that it is **categorical**
  * **Question.** Why is this an issue?
  * **Another Question.** How many cylinders are in cars, typically?
  
<br>
  
```{r warning=FALSE, message=FALSE, cache=TRUE}
data(mtcars)
ggplot(mtcars, aes(x = cyl,        # Number of cylinders
                   y = mpg)) +     # Miles per gallon 
  geom_point()
```

<br>

**What the heck?** Where are all the 5- and 7-cylinder cars?

  * Right, the 32 cars in dataset `mtcars` only have 4-, 6-, and 8-cylinder engines
  * R hasn't been told that variable `cyl` ("cylinders") is *categorical*
  * Let's try again with function `as.factor()`
  
<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
data(mtcars)
ggplot(mtcars, aes(x = as.factor(cyl),     # Cylinders "coerced" (changed) to factor
                   y = mpg)) +
  geom_point()
```

<br>

**Better.** But we can add some random noise by using the **geometry** `geom_jitter()`:

  * Argument `width =` and `height =` controls degree of random noise

<br>

```{r warning=FALSE, message=FALSE, cache=TRUE}
data(mtcars)
ggplot(mtcars, aes(x = as.factor(cyl),     # Still a factor
                   y = mpg)) +
  geom_jitter(width = 0.05)                # Jittered slightly, but "mpg" still precise
```

<br>

Right.

<br>

# Preset Themes

Because we're cutting it close on time, let's cover premade themes quickly.

  * The **themes layer** largely concerns style and **non-data ink**
  * All **themes** functions begin with `theme_`
  * Function `theme()` allows for total customization
  * Function `theme_set()` allows for *global themes* and will apply to all plots
    - More on this at a later time, we just had to cover **geometries**
  * Simply add your preferred **theme** with the `+` operator
  * There are dozens of premade **themes**, and dozens more with package `ggthemes`
    - Type `theme_` in the console and scroll through the autocomplete dropdown

<br>

Try `theme_light()`:

```{r warning=FALSE, message=FALSE, cache=TRUE}
library(ggthemes)
ggplot(diamonds,
       aes(x = cut,
           y = price)) +
  geom_jitter(width = 0.25,
              alpha = 0.05,
              color = "tomato") +
  theme_light()                       # Just add the theme layer onto the graph
```

<br>

Or `theme_fivethirtyeight()`:

```{r warning=FALSE, message=FALSE, cache=TRUE}
library(ggthemes)
ggplot(diamonds,
       aes(x = cut,
           y = price)) +
  geom_jitter(width = 0.25,
              alpha = 0.1,
              color = "tomato") +
  theme_fivethirtyeight()
```

<br>

Or `theme_minimal()`:

```{r warning=FALSE, message=FALSE, cache=TRUE}
library(ggthemes)
ggplot(diamonds,
       aes(x = cut,
           y = price)) +
  geom_jitter(width = 0.25,
              alpha = 0.05,
              color = "tomato") +
  theme_classic()
```

<br>

# Practice: Labeling & Themes

**Instructions.** Take any of the above plots and use the `+` operator to add function `labs()`.

  * In `labs()`, you can set graph labels for the following arguments:
    - `title =`
    - `subtitle =`
    - `caption =`
    - `x =`
    - `y =`
    - Make sure to put your text in quotations!
  * Select any dataset, between `mtcars`, `diamonds`, `economics`, or `iris`
  * Use function `names()` with the dataset name inside `()` to get the variable names
    - Alternatively, use function `help()` with the dataset for definitions
  * Select data ink for function `aes()`, a geometry, and non-data ink inside `geom_*()`
    - Alternatively, reuse an existing plot for convenience
  * Browse and select a theme you like using any `theme_*()` function
  * Make sure your data and non-data ink convey your finding clearly; polish appropriately

<br>

Thanks, everyone.