---
title: 'Intro to R: Dates & Times'
author: "Jamison R. Crawford, MPA"
date: "December 21, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
---

```{r setup, cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

**Latest Versions & Updates:** This markdown document was built using the following versions of *R* and *RStudio*:
 
* R v. 3.5.1
* RStudio v. 1.1.456
* Document v. 1.0
* Last Updated: 2018-12-20

# Review: Text Data

The following reviews some imporant concepts for working with text data, as seen in [Intro to R: Text Data](http://rpubs.com/JamisonCrawford/text):

**Classes:** Most variables are of a particular **class**, e.g. "numeric", "integer", "character", "logical", etc.

* Determine the class of a single variable using function `class()`
* Determine multiple variable classes using function `str()` on a data frame
* Confirm class via `TRUE` or `FALSE` with `is.*()` functions, e.g. `is.logical()`
* Coerce variables to a different class with `as.*()` functins, e.g. `as.character()`

**Character & Factor Classes:** The most common types of qualitative variables are of class **character** and **factor**.

* Character data are comprised of sequences of characters wrapped in quotes (`""`)
* Factors are defined explicitly as categorical, nominal, or discrete variables (all synonymous)
* Coerce variables to class "character" using `as.character()`, confirm with `is.character()`
* Coerce variables to class "factor" using `as.factor()`, confirm with `is.factor()`
* Create new factor variables using function `factor()`
* Factors are critical in visualizing and modeling data, and their categories are known as **levels**
* Factors may be ordinal (ordered), made possible with the `levels =` argument or `levels()` function

**Fundamentals of Text:** There are several easy-to-use Base R functions for "character" data.

* Combine strings of text with function `paste()` and specify delimiting characters with argument `sep =`
* Coerce and format numeric data with functions `format()` and `formatC()`
    * Package `scales` has additional formatting options, e.g. `dollar()`
* Print "character" data without quotes or position identifiers using function `writeLines()`
    * Print the same unquoted data with position identifiers using function `noquote()`

**Package "stringr" Functions:** The "stringr" package has an intuitive, unified framework common to the *Tidyverse*.

* Many critical "stringr" functions have a Base R counterpart but are easier to use
* Paste strings of text together using `str_c()`
* Determine the number of characters in a string using `str_length()`
* Extract substrings using `str_sub()`
* Return `TRUE` and `FALSE` values for pattern matches with `str_detect()`
* Return original values using `str_subset()`
* Return the frequency of pattern matches using `str_count()`
* Split strings into substrings with a specified delimiter using `str_split()`
* Find and replace specified patterns using `str_replace()` and `str_replace_all()`
* Trim whitespace on either or both sides of values using `str_trim()`
* "Pad" values with specified characters, e.g. leading zeroes, using `str_pad()`

<br>

# Dates & Times in R

**Overview:** We can easily create strings of text to represent dates and times.

* The character string "12/21/2018" may be human-readable, but not machine-readable
* One must coerce the class of these strings to be recognizable by R as dates or times
* Class **"POSIX"** (Portable Operating System Interface for Unix) is specifically for dates and datetimes
* Under the hood, class "POSIX" variables are stored as numbers, but format to be human-readable
    * These numbers represent either various numbers in date units or total seconds since Thursday, January 1, 1970
    * If a positive number, the date/datetime is *after* January 1, 1970
    * If a negative number, the date/datetime is *before* January 1, 1970
    * Seemingly arbitrary, this date actually derives from "Unix Time", a.k.a. ["The Epoch"](https://en.wikipedia.org/wiki/Unix_time)

<br>

## Class "POSIXlt" & "POSIXct"

**POSIXlt** stores datetimes in days, months, years, hours, minutes, and seconds relative to January 1, 1970.

Here, we'll use the current date and time using function `Sys.time()`:

```{r}
Sys.time()
my_date_lt <- as.POSIXlt(Sys.time())
unclass(my_date_lt)
```

<br>

**POSIXct** only stores datetimes in seconds relative to January 1, 1970.

```{r}
Sys.time()
my_date_ct <- as.POSIXct(Sys.time())
unclass(my_date_ct)
```

<br>

### Formatting "POSIXlt" & "POSIXct"

"POSIXlt" values may be formatted in a variety of ways, e.g.:

* "December 21, 2018"
* "21 December 2018"
* "2018-12-21"
* "12/21/18"
* "12-21-2018"

<br>

**Coercing to "POSIXlt":** We can coerce variables to "POSIXlt" class using function `as.POSIXlt()`, e.g.:

```{r}
my_date <- "2018-12-21"
print(my_date)
class(my_date)

my_date <- as.POSIXlt("2018-12-21")
print(my_date)
class(my_date)
```

<br>

**Standardized Formats:** Of these ways, the preferable and standard formatting is "YYYY-MM-DD", e.g. "2018-12-21". Why?

```{r}
my_dates <- c("October 01, 2018", "December 01, 2019", "November 01, 2018", "September 01, 2019")
print(my_dates)

sort(my_dates)
```

Note that without converting to class "POSIXlt", these dates are arranged **naively** in alphabetical order.

<br>

**Another Example:** Here, we only use numbers to identify dates. Observe:

```{r}
my_dates <- c("10/01/18", "12/01/19", "11/01/18", "09/01/19")
print(my_dates)

sort(my_dates)
```

These are also arranged **naively**, and incorrectly, in numerical order (note that 2019 comes both before and after 2018).

<br>

**Standard Formatting in Action:** Here we can see how the standard format is impossible to disarrange:

```{r}
my_dates <- c("2018-10-01", "2019-12-01", "2018-11-01", "2019-09-01")
print(my_dates)

sort(my_dates)
```

Even if arranged **naively** and not coerced to "POSIXlt" class, these dates will invariably sort correctly.

* This is because the units of measurement have a hierarchy of unit magnitude
* Hence why much of the rest of the world formats dates as, e.g. "21 December 2018"
    * Despite the opposite direction, these follow a hierarchy of unit magnitude
    * It's like the metric system in that it *actually makes sense*

<br>

**Formatting in Base R:** It's not a bad idea to know how this is done, but it's a pain in the rear.

* Therefore, we'll learn how to format dates using the *Tidyverse* package `lubridate`
* In Base R, you have to manually identify the elements of non-standard dates to properly parse them:
    * `%Y` is a 4-digit year
    * `%y` is a 2-digit year
    * `%m` is a 2-digit month
    * `%d` is a 2-digit day of the month
    * `%A` is the weekday, e.g. "Wednesday"
    * `%B` is the month, e.g. "February"
    * `%b` is the abbreviated month, e.g. "Feb"
* What's more, in Base R, you have to manually identify elements of non-standard timestamps:
    * `%H` is hours as a decimal number
    * `%I` is hours in AM/PM format as a decimal number
    * `%M` is minutes as a decimal number
    * `%S` is seconds as a decimal number
    * `%T` is shorthand for standard format: %H:%M:%S
    * `%p` is the AM/PM indicator
    
The standard format for timestamps are "HH:MM:SS", i.e. hours, minutes, and seconds.

* The same logic applies: Decreasing order of units of magnitude
* Therefore, datetime objects are standardized as "YYYY-MM-DD HH:MM:SS"

<br>

## Package "lubridate"

**Package** `lubridate` allows us to easily parse just about any date with simple, intuitive functions:

<br>

### Ways of Old

**Formatting datetimes**, as mentioned, is a huge pain. Here, we'll try to format `12/20/18`

```{r}
my_datetime <- "12/21/18 08:30:00 AM"
print(my_datetime)

as.POSIXlt(my_datetime)
```

Here, we get an error, since the input data isn't in "standard unambiguous format". To format in Base R:

<br>

**Formatting in Base R**, we have to manually format our datetimes, since they aren't "standard unambiguous" format.

* In other words, we have to tell R exactly how the date and time are formatted

```{r}
my_datetime <- "12/21/18 08:30:00 AM"
print(my_datetime)

as.POSIXlt(x = my_datetime, 
           format = "%m/%d/%y %H:%M:%S %p")
```

<br>

### Making Things Easier

**Package** `lubridate` is able to detect dates and times in "unambiguous" format with very little specification.

First, let's install and load `lubridate`:

```{r echo=TRUE, message=FALSE, warning=FALSE}
if(!require(lubridate)){install.packages("lubridate")}
library(lubridate)
```

<br>

**Formatting the Lubridate Way:** We'll take the same datetime and format it with function `mdy_hms()`.

```{r}
my_datetime <- "12/21/18 08:30:00 AM"
print(my_datetime)

mdy_hms(my_datetime)
```

Way, way, way easier! By using `mdy_hms()`, we told R that the format approximates "MM-DD-YYYY HH:MM:SS".

<br>

**Other Formats:** We can do this with a variety of datetime formats. Behold:

```{r}
my_datetime <- "12/21/2018 08:30:00 AM"
mdy_hms(my_datetime)

my_datetime <- "December 21, 2018 8:30:00"
mdy_hms(my_datetime)

my_datetime <-  "12/21/18 8:30:00"
mdy_hms(my_datetime)

my_datetime <- "Dec. 21, 2018 8:30:00"
mdy_hms(my_datetime)

my_datetime <- "Dec 21 18 8:30:00"
mdy_hms(my_datetime)

my_datetime <- "122118 083000"
mdy_hms(my_datetime)
```

We can get increasingly lazy with our dates, but `lubridate` gets the job done without fail.

<br>

**Similar Functions:** Other `lubridate` functions for formatting non-standard, unambiguous datetimes include:

* `mdy_hm()`
* `mdy_h()`
* `mdy()`
* `dmy_hms()`
* `dmy_hm()`
* `dmy_h()`
* `dmy`
* `ydm_hms()`
* `ydm_hm()`
* `ydm_h()`
* `ydm()`
* `ymd_hms()`
* `ymd_hm()`
* `ymd_h()`
* `ymd()`

Just remember: 

* `y` is "year"
* `m` is "month"
* `d` is "day"
* `h` is "hour"
* `m` is "minute"
* `s` is "second"

And you're good to go!

<br>

### Other Cool Functions

**Package** `lubridate` is replete with date-, time-, and datetime-related functions. Here's a few cool ones:

<br>

**Dates:** `date()` and `as_date()` are less intimidating and more human-readable ways to create class `POSIX` variables.

* Enter a new class, class **"Date"**, but this is the exact same thing as a "POSIClt" variable

```{r}
my_date <- "2018-12-21"
class(my_date)

my_date <- as_date(my_date)
class(my_date)
```

<br>

**Extracting Units of Time:** You can extract a unit of time from a datetime object, including:

* `year()`
* `month()`
* `week()`
* `day()`
* `weekdays()`
* `hour()`
* `minute()`
* `second()`
* `am()`
* `pm()`

Observe:

```{r}
my_datetime <- mdy_hms("December 21, 2018 08:30:00 AM")
year(my_datetime)
month(my_datetime)
week(my_datetime)
day(my_datetime)
weekdays(my_datetime)
hour(my_datetime)
minute(my_datetime)
second(my_datetime)
am(my_datetime)
pm(my_datetime)
```

<br>

**Timestamping:** Print current date and time with `now()`, similar to `Sys.time()` but easier:

```{r}
now()
```

<br>

**Determine Differences in Times:** Find time and date differences using simple arithmetic:

```{r eval = FALSE}
session_start <- mdy_hms("December 21, 2018 08:30:00 AM")
current_time <- now()

session_start - current_time
```

<br>

**Determine Difference in Dates:** Again, since dates and times are numbers under the hood, it's arithmetic!

```{r}
birthday <- mdy("August 10, 1989")
age <- today() - birthday

print(age)
class(age)
```

Storing this creates a new variable class: **"difftime"**.

* Class "difftime" objects may be converted to different units using Base R function `difftime()`

<br>

**Round Dates:** You can make dates uniform using `round_date()`, `floor_date()`, and `ceiling_date()`.

* You must specify the unit of time to which each function rounds
* Particularly useful for, e.g., grouping number of crimes by month

```{r}
crimes <- data.frame(crime = c("Larceny", "Larceny", "Arson"),
                     date = c("2018-04-03", "2018-04-19", "2018-05-11"))
crimes$date <- ymd(crimes$date)
print(crimes)

crimes$date <- floor_date(x = crimes$date, unit = "month")
print(crimes)
```

<br>

# Applied Practice

**Instructions:** Run the following code to read in housing code violation data from *Syracuse Open Data*.

* If you'd like, check out the documentation [here](http://data.syrgov.net/datasets/code-violations)
* The dataframe will be made available in your local environment: `violations`

```{r warning=FALSE, message=FALSE}
if(!require(readr)){install.packages("readr")}; library(readr)

url <- "https://opendata.arcgis.com/datasets/fb7233117df1443081541f220327f178_0.csv"
types <- "--c---T---c---c---------"
violations <- read_csv(file = url, col_types = types)
```

<br>

**Challenge 1:** Determine the class of variable `violation_date`. Convert it to class "Date".

<br>

**Challenge 2:** When was the earliest violation in `violations`? The latest?

* Hint: Consider using functions `min()` and `max()`

<br>

**Challenge 3:** What is the duration between the earliest and latest violation in `violations`?

<br>

**Challenge 4:** How long ago was the most recent violation as of today?

* Hint: Consider using functions `now()`, `max()`, and basic arithmetic

<br>

**Challenge 5:** Which month had the most reported violations according to the data?

* Hint #1: Consider using `floor_date()` to make instances uniform
* Hint #2: Consider using `table()`, only inputting the dataset and variable name with `$` notation
