---
title: 'Visualization with ggplot2: Part I'
author: "Jamison R. Crawford, MPA"
date: "April 12, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---

<style>
body {
text-align: justify}
</style>

<br>

# Review

For an introduction to data visualization in R, see [Visualization in Base R](http://rpubs.com/JamisonCrawford/graphics).

<br>

# The Grammar of Graphics

Before introducing package `ggplot2`, it's important to understand the theory on which it's built.

<br>

## Human Grammar

**Consider human language.** Native speakers rarely understand grammatical rules, but they use them intuitively.

For example, we may not know why the indefinite article is sometimes "a" and at other times "an". Observe:

* "There's a snake in my boot."
* "There's an elephant in the room."

We may also intuit the definite article, "the", in the case of superlatives. Observe:

* "Be better."
* "Be *the* best."

<br>

**Every Grammatical Element is Significant.** We use *parts of speech* like comparatives, superlatives, prepositions, definite and indefinite articles, verbs, nouns, adjectives, adverbs, phrasal verbs, and all other variety of grammatical elemets to create novel sentences. Observe a classic example:

> "The quick brown fox jumps over the lazy dog."

Note that if we change a single word, we change the signficance of the sentence.

> "The slow brown fox jumps over the lazy dog."

By using an antonym for "quick", the sentence is even more absurd. Let's make two more changes.

> "The slow brown fox runs over the ethereal dog."

Now we have a fox that can presumably operate a motor vehicle, and we delve into metaphysics a bit.

<br>

**We intuit the grammar of visualization.** But we often don't know the "rules" - if there are any.

* We know what looks good
* We know what looks bad

Like human sentences, machine visualization can be elegant and efficient in conveying meaning. 

Or it can be a salad of visual elements that disrespects your audience's time.

<br>

## A Grammar of Graphics

In 1999, Leland Wilkinson published [The Grammar of Graphics](https://www.springer.com/us/book/9780387245447) and a theoretical framework for such a grammar.

<br>

**Layers.** Graphics are comprised of distinct **layers** of grammatical elements.

* **Layers** are like *parts of speech* (e.g. nouns, verbs, etc.)
* Like complete sentences, there are essential layers to make complete graphical expressions
* These essential layers include the:
    * **Data Layer** or what data you use
    * **Aesthetics Layer** or what variable you wish to graph
    * **Gemoetry Layer** or the shape or form by which to graph them your variables
    
<br>

**Typical Conversation in Grammatical Terms.** Note what layers are really being discussed:

<br>

**Data****:**

> "Are you pulling occupations from O*NET or BLS? We only need SOC-level."

<br>

**Aesthetics:**

> "Could you color code the data points by ethnicity?"

<br>

**Geometry:**

> "I'm trying to emphasize the increase in cases of EBLL levels over time."

<br>

**Coordinates:** 

> "Can we zoom in on just the household incomes that are less than $45,000?"

<br>

**Statistics:**

> "Let's add one of those squiggly lines to make that trend really stand out."

<br>

**Facets:** 

> "Can we show multiple graphs that are organized by country of origin?"

<br>

**Themes:**

> "We can only use colors that are in the company logo. Wait, what is that?"

> "My two weeks' notice."

<br>

**A Unified Framework.** Individually, each element is a building block. As a whole, the **Grammar of Graphics** provides a common language for visualization experts.

<br>

**Mapping** or **Aesthetic Mapping** is simply depicting a variable by using these elements.

<br>

# ggplot2

Package `ggplot2` is a popular, flexible, and powerful visualization extension for R.

* Authored by Hadley Wickham of *RStudio*
* Built as a "wrapper library" around R package `grid`
* A core [Tidyverse](https://www.tidyverse.org/) package that interfaces seamlessly within the Tidyverse ecosystem
* Implements many best practices by default, like the color decodability research of [Cynthia Brewer](https://en.wikipedia.org/wiki/Cynthia_Brewer)
* Works best with [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf)

Further resources:

* The [Tidyverse Overview](https://ggplot2.tidyverse.org/) of `ggplot2`
* The `ggplot2` [Vignette](https://www.rdocumentation.org/packages/ggplot2/versions/3.1.1)
* The `ggplot2` [Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
* General tips for `ggplot2` from the [R Graph Gallery](https://www.r-graph-gallery.com/portfolio/ggplot2-package/)
* [CRAN Documentation](https://cran.r-project.org/web/packages/ggplot2/index.html) for `ggplot2`
* The `ggplot2` [Extensions Gallery](http://www.ggplot2-exts.org/gallery/)

<br>

## Installing & Loading

Installing and loading `ggplot2` is easy.

<br>

Use function `install.packages()` to install `ggplot2`. 

* Make sure to use quotes around `ggplot2`
* You only have to install it once

```{r eval = FALSE}
install.packages("ggplot2") 
```

<br>

Use function `library()` to load `ggplot2`.

* You don't need quotes, but you can use them (both are shown)
* You must load `ggplot2` every time you start a new session

```{r}
library(ggplot2)
library("ggplot2")
```

<br>

## Loading Practice Data

We'll use the practice dataset, `diamonds`, which comes with the `ggplot2` package.

* The go-to practice dataset for `ggplot2`
* Contains ~54,000 observations, each a diamond
* Variables include `color`,, `price`, `clarity`, etc.
* Learn more by running `?diamonds` or `help(diamonds)`
* Load the dataset with function `data()`

```{r}
data(diamonds)
```

<br>

Some other functiosn for exploring the `diamonds` dataset include:

```{r}
str(diamonds)          # The structure of the data
names(diamonds)        # Variables names, or use...
colnames(diamonds)
dim(diamonds)          # Tabular dimensions 
```

<br>

## Functions & Additions

Package `ggplot2` has an absurd number of functions, but if you grasp the theory and practice, you'll get it.

Each function begins with a name corresponding to the **layers** identified above.

* Function `ggplot()` corresponds to the **Data Layer**
* Function `aes()` corresponds to the **Aesthetics Layer**
* Functions beginning with `geom_` correspond to the **Geometry Layer**
* Functions beginning with `theme_` correspond to the **Themes Layer**
* Functions beginning with `facet_`correspond to the **Facets Layer**
* Etc.

<br>

The **Addition Operator**, or `+`, connects all of these functions into a single visualization. 

For example, I've saved a **grob**, or **Graphical Object** that contains a plot, to demonstrate.

* You don't really need to know what a **grob** is, yet, so don't worry about that now
* Observe the use of the **Addition Operator** to modify **grobs** by adding new functions

```{r message=FALSE, echo=FALSE, warning=FALSE, cache=TRUE}
my_grob <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

```{r cache=TRUE}
my_grob
```

<br>

Note that we can add, for example, a new them with a `theme_*()` function.

* Here, the `*` represents anything that may follow `theme_`

```{r cache=TRUE}
my_grob +
  theme_classic()
```

<br>

Let's try a different premade theme - there are a ton!

```{r cache=TRUE}
my_grob +
  theme_minimal()
```

```{r cache=TRUE}
my_grob +
  theme_light()
```

<br>

**Package Extensions.** You can even add new themes with package `ggthemes`.

* Make sure to install with `install.packages()` and load with `library()`

```{r cache=TRUE, warning=FALSE, message=FALSE}
library(ggthemes)
my_grob +
  theme_fivethirtyeight()
```

Pretty neat. *There's even a color scheme for each of Wes Anderson's movies.*

We'll discuss extensions at a later session.

<br>

## Intializing a Plot

The anatomy of a plot is simple. As mentioned, three **layers** are essential:

* The **Data Layer** with function `ggplot()` and the name of the dataset
* The **Aesthetics Layer** with function `aes()` and the variables to **map**
* The **Geomtry Layer** with functions starting with `geom_` for the shape of the plot

<br>

You can chain all of these together with the **Addition Operator**.

* Here, we map variables to the x-axis and y-axis using arguments `x =` and `y =` in `aes()`
* Note that you do not need to reference the dataset again, once called in function `ggplot()`
* What's more, you don't have to put variable names in quotes, like `carat` and `price`
    * This is called a **Quoting Function**

```{r cache = TRUE}
ggplot(data = diamonds) +
  aes(x = carat, y = price) +
  geom_point()
```

<br>

**The "Right" Way.** Although it's much easier and cleaner to keep these functions separate:

* There is a preferred syntax that combines functions `ggplot()` and `aes()`
* This isn't immediately useful but is paramount for advanced visualizations
* Why? You can use different datasets in the same graphic!
* If you're new to coding, do whatever is easier for now

```{r cache=TRUE}
ggplot(data = diamonds, aes(x = carat, 
                            y = price)) +
  geom_point()
```

<br>

## Aesthetics v. Attributes

The difference between **Attributes** and **Aesthetics** is extremely important but very simple.

**Aesthetic Mappings** depict a variable from your data

* That is, `x =`, `y =`, `color =`, `fill =`, `alpha =`, etc. *all represent data*
* This helps with depicting multiple variables in a single graphic
* **The location of aesthetics** belong in function `aes()`
    * That is, the **Aesthetics Layer**

<br>

**Attributes** do not depict data - this is called **Non-Data Ink**

* Almost the same arguments apply, like `color =` and `fill =`
* This helps with decoding a visualization and facilitating understanding
* **The location of attributes** belong in any functiosn starting with `geom_`
    * That is, the **Geometry Layer**
    
<br>

### Examples

Let's check out the same plot with an extra **Aesthetic Mapping**. Here, we'll add a new variable, each diamond's quality of `cut`, and map it to `color =` in function `aes()`:

```{r cache=TRUE}
ggplot(data = diamonds, aes(x = carat, 
                            y = price,
                            color = cut)) +
  geom_point()
```

We can already gain new insights by mapping the variable `cut`.

**Question.** What can you tell about the relationship between `price` and `cut`?

<br>

Now, observe an example of an **Attribute** or **Non-Data Ink**. Because we're not depicting a new variable, we don't put these arguments in functon `aes()`. Instead, we put them in function `geom_point()`, and we can choose the **Attributes** that help us to interpret the plot. Specifically:

* `alpha =` sets the transparency of data points; helpful because of heavy overlap
* `color =` helps us discern transparent points from the default grey background of `ggplot2`
* We can use `theme_light()` or other functions for a non-grey background, if needed

```{r cache=TRUE}
ggplot(data = diamonds, aes(x = carat, 
                            y = price)) +
  geom_point(alpha = 0.1,
             color = "tomato") +
  theme_light()
```

We haven't added any new data to the plot, but now we can more easily interpret it.

**Question.** What insights are made more clear by the parsimonious use of **Non-Data Ink**?

<br>

# Practice

I'm went over my hourage cap for session prep like three hours ago, and I have some rare birds to hunt in *Red Dead Redemption 2*, but this practice may still be useful.

<br>

**Instructions.** Using the same plot with which we've practiced, experiment with:

* Different **Aesthetic Mappings** for each variable
* Remember to put them inside function `aes()` because they are **Data Ink**
* Run the following to get their names:

```{r}
names(diamonds)
```

<br>

Also, try:

* Different **Attributes** to make the plot more appealing and decodable
* Remember to put them inside function `geom_point()` because they are **Non-Data Ink**

<br>

It's dangerous to go alone. Take this:

* [Colors in R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)
* [Aesthetic Specifications](https://cran.r-project.org/web/packages/ggplot2/vignettes/ggplot2-specs.html), like `shape =`, `size =`, `alpha =`, etc.

<br>

Thanks for reading!